---
title: "Building LR models for progression sets using Clinical data alone"
output: html_notebook
---

Load clinical data and progression annotations
```{r}
load("../../data/clinical_data.RData")
datasets <- clinical_out
load("../../data/progressionIDs052017_BRCA_withData.RData")
progressed_BRCA <- progressed
nonProgressed_BRCA <- nonProgressed
load("../../data/progressionIDs052017_PRAD_withData.RData")
progressed_PRAD <- progressed
nonProgressed_PRAD <- nonProgressed
load("../../data/progressionIDs052017_KIRC_withData.RData")
progressed_KIRC <- progressed
nonProgressed_KIRC <- nonProgressed
load("../../data/progressionIDs052017_LUSC_withData.RData")
progressed_LUSC <- progressed
nonProgressed_LUSC <- nonProgressed
rm(list = c("progressed", "nonProgressed", "clinical_out"))
```
A function to check if NA-valued coefficients are present in the GLM model (that were failed to estimate due to lack of evidence), and if that is the case, hack the GLM model and provide neutral coefficient value and rescale the other factors accordingly
```{r}
fixNAcoefficients <- function(glmModel) {
  factor_vars <- names(glmModel$xlevels)
  touchedFactors <- FALSE
  for (factor_var in factor_vars) {
    xlevels <- grep(pattern = factor_var, x = names(coef(glmModel)), value = TRUE)
    if (!is.na(summary(is.na(glmModel$coefficients[xlevels]))["TRUE"])) {
      touchedFactors <- TRUE
  	  if (length(xlevels) > 1) {
        glmModel$coefficients["(Intercept)"] <- glmModel$coefficients["(Intercept)"] + mean(glmModel$coefficients[xlevels], na.rm=TRUE)
        glmModel$coefficients[xlevels][is.na(glmModel$coefficients[xlevels])] <- mean(glmModel$coefficients[xlevels], na.rm = TRUE)
        glmModel$coefficients[xlevels] <- scale(glmModel$coefficients[xlevels], center=TRUE, scale=FALSE)[,1]
  	  } else glmModel$coefficients[xlevels] <- 0
    }
  }
  return(glmModel)
}
```
A function to hack GLM model and check if missing factors are present (with respect to new data), and if that is the case, provide neutral coefficient value for the missing factors and calculate the resulting 'response' value
```{r}
fixMissingFactor <- function(factor_var, factor, glmModel) {
  # Following procedure adapted from proposal made by Sam Thomas:
  # https://stackoverflow.com/questions/31917878/handling-unknown-factor-levels-in-r-glm
  # coefficients from model
  coefficients <- coef(glmModel)[grepl(factor_var, names(coef(glmModel)))]
  # calculate coefficient for new factors
  coefficients_offset <- mean(coefficients, na.rm=TRUE)
  # insert missing coefficient
  coefficients[factor] <- coefficients_offset
  # center coefficients
  coefficients <- scale(coefficients, center=TRUE, scale=FALSE)[, 1]
  
  # replace coefficients from model
  modelcoef <- coef(glmModel)
  # add offset to intercept
  modelcoef[["(Intercept)"]] <- modelcoef[["(Intercept)"]] + coefficients_offset
  # all new coefficients
  modelcoef[names(coefficients)] <- coefficients
  glmModel$coefficients <- modelcoef
  return(glmModel)
}

predict_whileCheckingMissingFactors <- function (glmModel, newdata, type="response") {
  
  previous_na_action <- options('na.action')
  options(na.action='na.pass')
  newxreg <- model.matrix(glmModel$formula, data=newdata)
  options(na.action=previous_na_action$na.action)
  newxreg[is.na(newxreg)] <- 0
  
  predictions <- rep(NA, nrow(newdata))
  factor_vars <- names(glmModel$xlevels)
  missing_factors <- setdiff(colnames(newxreg), names(glmModel$coefficients))
  touchedFactors <- FALSE
  
  if (length(missing_factors) > 0) {
    touchedFactors <- TRUE
    factor_vars_indexes <- NULL
    for (i in 1:length(missing_factors)) {
      missing_factor <- missing_factors[i]
      tmp <- unlist(lapply(1:length(factor_vars), FUN = function(x) if (length(grep(factor_vars[x], missing_factor)) > 0) x ))
      factor_vars_indexes[[factor_vars[tmp]]] <- tmp
      glmModel <- fixMissingFactor(factor_var=factor_vars[tmp], factor=missing_factor, glmModel)
    }
    rm(tmp)
  }
  glmModel <- fixNAcoefficients(glmModel)
  # Beta0 + Beta1x...
  #glmModel$coefficients[["(Intercept)"]] + 
  pcoef <- newxreg[,] %*% glmModel$coefficients[colnames(newxreg[,])]
  #predicted response
  if (type == "response") {
    require(Brobdingnag) # using library for precision when large and small numbers go into formula
    predictions <- unlist(lapply(pcoef, function(x) {x <- as.brob(x); as.numeric(exp(x) / (1 + exp(x)))}))
  } else (print("Warning, only 'response' type prediction was implemented"))
  return(predictions)
}
```

BRCA data
```{r}
library(pROC)
library(dplyr)

clinical_data_BRCA <- datasets[["BRCA"]]@Clinical[c(progressed_BRCA, nonProgressed_BRCA),]
clinical_data_BRCA$progressed <- rownames(clinical_data_BRCA) %in% progressed_BRCA
factors <- c("pathologic_stage", "pathology_T_stage", "pathology_N_stage", "pathology_M_stage", "histological_type", "race", "ethnicity")
for (i in 1:length(factors)) {
  clinical_data_BRCA[,factors[i]] <- factor(clinical_data_BRCA[,factors[i]])
}

nfolds <- 10
# Use split to generate CV-sets
set.seed(1) # Use the same sets accross genes
G1_cv_sets <- 1:length(progressed_BRCA) %>% sample %>% split(1:nfolds)
G2_cv_sets <- 1:length(nonProgressed_BRCA) %>% sample %>% split(1:nfolds)

predictions <- NULL
for (fold in 1:nfolds) {
  # subset the groups for current fold
	G1_train <- progressed_BRCA[-G1_cv_sets[[fold]]]
	G2_train <- nonProgressed_BRCA[-G2_cv_sets[[fold]]]	
  G1_eval <- progressed_BRCA[G1_cv_sets[[fold]]]
	G2_eval <- nonProgressed_BRCA[G2_cv_sets[[fold]]]
	
	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + histological_type + as.numeric(number_of_lymph_nodes) + race + ethnicity, data = clinical_data_BRCA[c(G1_train, G2_train),], family = "binomial", maxit=100,  na.action = "na.exclude")
	
	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_BRCA[c(G1_eval, G2_eval),], type="response")
  #model <- fixNAcoefficients(model)
  #predictions[c(G1_eval, G2_eval)] <- predict(model, newdata = clinical_data_BRCA[c(G1_eval, G2_eval),], type="response")
}
auc(response=clinical_data_BRCA$progressed, predictor=predictions)
```

PRAD data
```{r}
require(pROC)
require(dplyr)

clinical_data_PRAD <- datasets[["PRAD"]]@Clinical[c(progressed_PRAD, nonProgressed_PRAD),]
clinical_data_PRAD$progressed <- rownames(clinical_data_PRAD) %in% progressed_PRAD
factors <- c("pathology_T_stage", "pathology_N_stage", "radiation_therapy", "histological_type", "residual_tumor", "race", "ethnicity") # no differential race or ethnicity was observed in the cohort
for (i in 1:length(factors)) {
  clinical_data_PRAD[,factors[i]] <- factor(clinical_data_PRAD[,factors[i]])
}

nfolds <- 10
# Use split to generate CV-sets
set.seed(1) # Use the same sets accross genes
G1_cv_sets <- 1:length(progressed_PRAD) %>% sample %>% split(1:nfolds)
G2_cv_sets <- 1:length(nonProgressed_PRAD) %>% sample %>% split(1:nfolds)

predictions <- NULL
for (fold in 1:nfolds) {
  # subset the groups for current fold
	G1_train <- progressed_PRAD[-G1_cv_sets[[fold]]]
	G2_train <- nonProgressed_PRAD[-G2_cv_sets[[fold]]]	
  G1_eval <- progressed_PRAD[G1_cv_sets[[fold]]]
	G2_eval <- nonProgressed_PRAD[G2_cv_sets[[fold]]]
	
	model <- glm(progressed ~ as.numeric(years_to_birth) + pathology_T_stage + pathology_N_stage + radiation_therapy + histological_type + residual_tumor + as.numeric(number_of_lymph_nodes), data = clinical_data_PRAD[c(G1_train, G2_train),], family = "binomial", maxit=100,  na.action = "na.exclude") # no differential race or ethnicity was observed in the cohort
	
	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_PRAD[c(G1_eval, G2_eval),], type="response")
  #model <- fixNAcoefficients(model)
  #predictions[c(G1_eval, G2_eval)] <- predict(model, newdata = clinical_data_PRAD[c(G1_eval, G2_eval),], type="response")
}
auc(response=clinical_data_PRAD$progressed, predictor=predictions)
```

KIRC data
```{r}
require(pROC)
require(dplyr)

clinical_data_KIRC <- datasets[["KIRC"]]@Clinical[c(progressed_KIRC, nonProgressed_KIRC),]
clinical_data_KIRC$progressed <- rownames(clinical_data_KIRC) %in% progressed_KIRC
factors <- c("pathologic_stage", "pathology_T_stage", "pathology_N_stage", "pathology_M_stage", "gender", "radiation_therapy", "karnofsky_performance_score", "histological_type", "race", "ethnicity")
for (i in 1:length(factors)) {
  clinical_data_KIRC[,factors[i]] <- factor(clinical_data_KIRC[,factors[i]])
}

nfolds <- 10
# Use split to generate CV-sets
set.seed(1) # Use the same sets accross genes
G1_cv_sets <- 1:length(progressed_KIRC) %>% sample %>% split(1:nfolds)
G2_cv_sets <- 1:length(nonProgressed_KIRC) %>% sample %>% split(1:nfolds)

predictions <- NULL
for (fold in 1:nfolds) {
  # subset the groups for current fold
	G1_train <- progressed_KIRC[-G1_cv_sets[[fold]]]
	G2_train <- nonProgressed_KIRC[-G2_cv_sets[[fold]]]	
  G1_eval <- progressed_KIRC[G1_cv_sets[[fold]]]
	G2_eval <- nonProgressed_KIRC[G2_cv_sets[[fold]]]
	
	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + gender + radiation_therapy + race + ethnicity, data = clinical_data_KIRC[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action = "na.exclude") # had to remove Karnofsky perf. score due to missingness
	
	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_KIRC[c(G1_eval, G2_eval),], type="response")
  #model <- fixNAcoefficients(model)
  #predictions[c(G1_eval, G2_eval)] <- predict(model, newdata = clinical_data_KIRC[c(G1_eval, G2_eval),], type="response")
}
auc(response=clinical_data_KIRC$progressed, predictor=predictions)
```

LUSC data
```{r}
require(pROC)
require(dplyr)

clinical_data_LUSC <- datasets[["LUSC"]]@Clinical[c(progressed_LUSC, nonProgressed_LUSC),]
clinical_data_LUSC$progressed <- rownames(clinical_data_LUSC) %in% progressed_LUSC
factors <- c("pathologic_stage", "pathology_T_stage", "pathology_N_stage", "pathology_M_stage", "gender", "radiation_therapy", "karnofsky_performance_score", "histological_type", "residual_tumor", "race", "ethnicity")
for (i in 1:length(factors)) {
  clinical_data_LUSC[,factors[i]] <- factor(clinical_data_LUSC[,factors[i]])
}

nfolds <- 10
# Use split to generate CV-sets
set.seed(1) # Use the same sets accross genes
G1_cv_sets <- 1:length(progressed_LUSC) %>% sample %>% split(1:nfolds)
G2_cv_sets <- 1:length(nonProgressed_LUSC) %>% sample %>% split(1:nfolds)

predictions <- NULL
for (fold in 1:nfolds) {
  # subset the groups for current fold
	G1_train <- progressed_LUSC[-G1_cv_sets[[fold]]]
	G2_train <- nonProgressed_LUSC[-G2_cv_sets[[fold]]]	
  G1_eval <- progressed_LUSC[G1_cv_sets[[fold]]]
	G2_eval <- nonProgressed_LUSC[G2_cv_sets[[fold]]]
	
	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + radiation_therapy + as.numeric(number_pack_years_smoked) + residual_tumor + gender + race, data = clinical_data_LUSC[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action = "na.exclude") # ethnicity and karnofsky variables were unreliable in estimation
	
	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_LUSC[c(G1_eval, G2_eval),], type="response")
  #model <- fixNAcoefficients(model)
  #predictions[c(G1_eval, G2_eval)] <- predict(model, newdata = clinical_data_LUSC[c(G1_eval, G2_eval),], type="response")
}
auc(response=clinical_data_LUSC$progressed, predictor=predictions)
```

Additional sensitivity analysis of BRCA data:
```{r}
library(pROC)
library(dplyr)

nfolds <- 1:10
aucs <- NULL
for (nfold in 1:length(nfolds)) {
  folds <- nfolds[nfold]
  # Use split to generate CV-sets
  set.seed(1) # Use the same sets accross genes
  G1_cv_sets <- 1:length(progressed_BRCA) %>% sample %>% split(1:folds)
  G2_cv_sets <- 1:length(nonProgressed_BRCA) %>% sample %>% split(1:folds)
  
  predictions <- NULL
  for (fold in 1:folds) {
    if (folds == 1) {
      G1_train <- G1_eval <- progressed_BRCA
      G2_train <- G2_eval <- nonProgressed_BRCA
    } else {
      # subset the groups for current fold
    	G1_train <- progressed_BRCA[-G1_cv_sets[[fold]]]
    	G2_train <- nonProgressed_BRCA[-G2_cv_sets[[fold]]]	
      G1_eval <- progressed_BRCA[G1_cv_sets[[fold]]]
    	G2_eval <- nonProgressed_BRCA[G2_cv_sets[[fold]]]
    }
    
  	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + histological_type + as.numeric(number_of_lymph_nodes) + race + ethnicity, data = clinical_data_BRCA[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action="na.exclude")
  	
  	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_BRCA[c(G1_eval, G2_eval),], type="response")
  	#model <- fixNAcoefficients(model)
  	#predictions[c(G1_eval, G2_eval)] <- predict(model, newData = clinical_data_BRCA[c(G1_eval, G2_eval),], type="response")
  }
  aucs[[nfold]] <- auc(response=clinical_data_BRCA$progressed, predictor=predictions)
}
aucs <- unlist(aucs)
plot(1:10, aucs)
lines(1:10, aucs)
```
Additional sensitivity analysis of PRAD data:
```{r}
require(pROC)
require(dplyr)

nfolds <- 1:10
aucs <- NULL
for (nfold in 1:length(nfolds)) {
  folds <- nfolds[nfold]
  # Use split to generate CV-sets
  set.seed(1) # Use the same sets accross genes
  G1_cv_sets <- 1:length(progressed_PRAD) %>% sample %>% split(1:folds)
  G2_cv_sets <- 1:length(nonProgressed_PRAD) %>% sample %>% split(1:folds)
  
  predictions <- NULL
  for (fold in 1:folds) {
    if (folds == 1) {
      G1_train <- G1_eval <- progressed_PRAD
      G2_train <- G2_eval <- nonProgressed_PRAD
    } else {
      # subset the groups for current fold
    	G1_train <- progressed_PRAD[-G1_cv_sets[[fold]]]
    	G2_train <- nonProgressed_PRAD[-G2_cv_sets[[fold]]]	
      G1_eval <- progressed_PRAD[G1_cv_sets[[fold]]]
    	G2_eval <- nonProgressed_PRAD[G2_cv_sets[[fold]]]
    }
    
  	model <- glm(progressed ~ as.numeric(years_to_birth) + pathology_T_stage + pathology_N_stage + radiation_therapy + histological_type + residual_tumor + as.numeric(number_of_lymph_nodes), data = clinical_data_PRAD[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action="na.exclude") # no differential race or ethnicity was observed in the cohort
  	# getting error at 3 folds: contrasts can be applied only to factors with 2 or more levels
  	
  	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_PRAD[c(G1_eval, G2_eval),], type="response")
  	#model <- fixNAcoefficients(model)
  	#predictions[c(G1_eval, G2_eval)] <- predict(model, newData = clinical_data_PRAD[c(G1_eval, G2_eval),], type="response")
  }
  aucs[[nfold]] <- auc(response=clinical_data_PRAD$progressed, predictor=predictions)
}
aucs <- unlist(aucs)
plot(1:10, aucs)
lines(1:10, aucs)
```
Additional sensitivity analysis of KIRC data:
```{r}
require(pROC)
require(dplyr)

nfolds <- 1:10
aucs <- NULL
for (nfold in 1:length(nfolds)) {
  folds <- nfolds[nfold]
  # Use split to generate CV-sets
  set.seed(1) # Use the same sets accross genes
  G1_cv_sets <- 1:length(progressed_KIRC) %>% sample %>% split(1:folds)
  G2_cv_sets <- 1:length(nonProgressed_KIRC) %>% sample %>% split(1:folds)
  
  predictions <- NULL
  for (fold in 1:folds) {
    if (folds == 1) {
      G1_train <- G1_eval <- progressed_KIRC
      G2_train <- G2_eval <- nonProgressed_KIRC
    } else {
      # subset the groups for current fold
    	G1_train <- progressed_KIRC[-G1_cv_sets[[fold]]]
    	G2_train <- nonProgressed_KIRC[-G2_cv_sets[[fold]]]	
      G1_eval <- progressed_KIRC[G1_cv_sets[[fold]]]
    	G2_eval <- nonProgressed_KIRC[G2_cv_sets[[fold]]]
    }
    
  	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + gender + radiation_therapy + race + ethnicity, data = clinical_data_KIRC[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action="na.exclude") # had to remove Karonofsky perf. score due to missingness
  	# getting error at 2 folds: contrasts can be applied only to factors with 2 or more levels
  	
  	
  	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_KIRC[c(G1_eval, G2_eval),], type="response")
  	#model <- fixNAcoefficients(model)
  	#predictions[c(G1_eval, G2_eval)] <- predict(model, newData = clinical_data_KIRC[c(G1_eval, G2_eval),], type="response")
  }
  aucs[[nfold]] <- auc(response=clinical_data_KIRC$progressed, predictor=predictions)
}
aucs <- unlist(aucs)
plot(1:10, aucs)
lines(1:10, aucs)
```
Additional sensitivity analysis of LUSC data:
```{r}
nfolds <- 1:10
aucs <- NULL
for (nfold in 1:length(nfolds)) {
  folds <- nfolds[nfold]
  # Use split to generate CV-sets
  set.seed(1) # Use the same sets accross genes
  G1_cv_sets <- 1:length(progressed_LUSC) %>% sample %>% split(1:folds)
  G2_cv_sets <- 1:length(nonProgressed_LUSC) %>% sample %>% split(1:folds)
  
  predictions <- NULL
  for (fold in 1:folds) {
    if (folds == 1) {
      G1_train <- G1_eval <- progressed_LUSC
      G2_train <- G2_eval <- nonProgressed_LUSC
    } else {
      # subset the groups for current fold
    	G1_train <- progressed_LUSC[-G1_cv_sets[[fold]]]
    	G2_train <- nonProgressed_LUSC[-G2_cv_sets[[fold]]]	
      G1_eval <- progressed_LUSC[G1_cv_sets[[fold]]]
    	G2_eval <- nonProgressed_LUSC[G2_cv_sets[[fold]]]
    }
    
  	model <- glm(progressed ~ as.numeric(years_to_birth) + pathologic_stage + pathology_T_stage + pathology_N_stage + pathology_M_stage + radiation_therapy + as.numeric(number_pack_years_smoked) + residual_tumor + gender + race, data = clinical_data_LUSC[c(G1_train, G2_train),], family = "binomial", maxit=100, na.action="na.exclude") # ethnicity and karnofsky variables were unreliable in estimation
  	
  	predictions[c(G1_eval, G2_eval)] <- predict_whileCheckingMissingFactors(model, newdata = clinical_data_LUSC[c(G1_eval, G2_eval),], type="response")
  	#model <- fixNAcoefficients(model)
  	#predictions[c(G1_eval, G2_eval)] <- predict(model, newData = clinical_data_LUSC[c(G1_eval, G2_eval),], type="response")
  }
  aucs[[nfold]] <- auc(response=clinical_data_LUSC$progressed, predictor=predictions)
}
aucs <- unlist(aucs)
plot(1:10, aucs)
lines(1:10, aucs)
```

